Solution-aware vs global ReLU selection: \\ partial MILP strikes back for DNN verification

### 2025-01-27 CAV 2025 TODO/Comparison.tex
comparison

First, we adapt from the BaB-SR \cite{BaB} and FSB \cite{FSB} functions (replacing the symbols from the original paper with those used in our paper.), selecting a node to branch on for BaB, global ranking (GR) functions to select nodes for pMILP. Intuitively, we extract the scoring $s_{SR}$ and $s_{FSB}$ from both BaB-SR and FSB, as the BaB bounding step is not adapted for pMILP. They both work by backpropagating gradients vectors $\bm{\lambda}$ from the neurons under consideration in layer $n$, back to neurons to be potentially selected. To do so, they consider the rate of $\ReLU(\bm{u}_k)$ to be $r(\bm{u}_k)=\frac{\UB(\bm{u}_k)}{\UB(\bm{u}_k)-\LB(\bm{u}_k)}$.

\begin{align*}
\bm{\lambda}_{n-1} = -{(\bm{W}^n)}^T\bm{1}, \hspace*{4ex}  	\bm{\lambda}_{k-1} = {(\bm{W}^k)}^T\big(\frac{\max(0,\UB(\bm{u}_k))}{\max(0,\UB(\bm{u}_k))-\min(0,\LB(\bm{u}_k))}\odot\bm{\lambda}_{k}\big) \hspace*{4ex}  k\in [n-1,2]
\end{align*}

@@

Consider the example in Fig. \ref{img:FSB_example}.
We have $\lambda(b)=d$, $\lambda(a)=\frac{cd}{2}$.
Let us do a case analysis:
The most favorable cases are $c < 0 < d$ and $d < 0 < c$: 
as $cd <0$, we have $s_{FSB}(a)=\max(0,\frac{cd}{4}) = 0$.
Now, $a$ and $\hat{a}$ need to be minimized in order to maximize $z$, because $cd <0$. 
For the triangle approximation, having a value $sol(a)$ for $a$, the minimal value for $\hat{a}$ is $\ReLU(sol(a))$. If we open $a$, then the value for $\hat{a}$ will also be $\ReLU(sol(a))$. That is, opening this ReLU is not helping in the approximation of $z$, as correctly predicted by the score.

Case $c>0,d>0$: $s_{FSB}(a)=\frac{cd}{4}$.
Indeed, the value of $a,\hat{a},b,\hat{b}$ should be maximized to maximize the value of $z$, because $W_{bz}=d>0$ and $W_{ab}=c>0$. Hence each variation of $b$ by $\Delta(b)$ will have an impact $r(b) \cdot \Delta(b)$.
Now, let us call $sol(a)$ the maximum value for $a$.
Then the maximum value for $\hat{a}$ is $r(a)\cdot (sol(a)-LB(a))$ according to the triangle approximation when $\ReLU(a)$ is not opened. Now, if $\ReLU(a)$ is opened, then 
the value of $\hat{a}$ is $\ReLU(sol(a))$: the difference is at most $r(a) \cdot LB(a) = \frac{1}{2}$. Going through $b$, the difference is at most $\frac{c}{2}$, which means a difference at most $\frac{c}{4}$ for $\hat(b)$, considering the upper bound of the triangle approximation, meaning a difference at most $\frac{cd}{4}$, which is what is computed by 
$s_{FSB}(a)$. Of course, the difference of opening $\ReLU(a)$ could also be 0 e.g. if $\sol(a)=\LB(a)$ or $\sol(a)=\UB(a)$.

The last case is the most problematic: 
Case $c<0,d<0$, implying that $s_{FSB}(a)=\frac{cd}{4}$.

\begin{figure}[b!]
	\centering
	\begin{tikzpicture}

@@

		\draw[->,thick] (input1) -- (hidden1) node[near start, above] {$c$};
		\draw[->,thick] (input1) -- (hidden2)node[near start, below] {$-10$};

@@

		\draw[->, thick] (hidden3) -- (hidden5) node[near start, above] {$1$};

@@

		\draw[->,thick] (output1) -- (output3)node[near start, above] {$d$};
		\draw[->,thick] (output2) -- (output3)node[near start, below] {$1$};

	\end{tikzpicture}
	\caption{}
	\label{img:FSB_example}
\end{figure}

\begin{figure}[t!]
	\begin{tikzpicture}[scale=1, >=stealth]

		% Draw ReLU function
		\draw[line width=0.4mm, blue] (-3,0) -- (0,0);
		\draw[thick, blue] (0,0) -- (2.5,2.5) node[below, shift={(-1.1,-1.9)}] {$\hat{b} = \ReLU(b)$};
		\draw[thick, blue] (-3,0) -- (2.5,2.5) node[above, shift={(-6.5,-2)}] {$\hat{b} = \frac{\UB}{\UB-\LB} (b-\LB)$};

@@

	\end{tikzpicture}
		\caption{Triangle approximation}
	\label{img:FSB}
	\end{figure}

\begin{figure}[t!]
\begin{tikzpicture}[scale=1, >=stealth]

@@

\end{tikzpicture}
	\caption{}
\label{img:Utility2}
\end{figure}

\subsection*{An example when FSB is not accurate}

@@

The plot shows that our Utility function is significantly better than FSB, and other methods.

>>>>

First, we adapt from the BaB-SR \cite{BaB} and FSB \cite{FSB} functions (replacing the symbols from the original paper with those used in our paper.), selecting a node to branch on for BaB, global ranking (GR) functions to select nodes for pMILP. Intuitively, we extract the scoring $s_{SR}$ and $s_{FSB}$ from both BaB-SR and FSB, as the BaB bounding step is not adapted for pMILP. They both work by backpropagating gradients vectors $\bm{\lambda}$ from the neurons under consideration in layer $n$, back to neurons to be potentially selected. To do so, they consider the rate of $\ReLU(\bm{u}_k)$ to be 
$r(\bm{u}_k)=\frac{\max(0,\UB(\bm{u}_k))}{\max(0,\UB(\bm{u}_k))-\min(0,\LB(\bm{u}_k))} \in [0,1]$.

\begin{align*}
\bm{\lambda}_{n-1} = -{(\bm{W}^n)}^T\bm{1}, \hspace*{4ex}  	\bm{\lambda}_{k-1} = {(\bm{W}^k)}^T\big( r(\bm{u}_k) \odot\bm{\lambda}_{k}\big) \hspace*{4ex}  k\in [n-1,2]
\end{align*}

@@

\begin{figure}[t!]
	\centering
	\begin{tikzpicture}

@@

		\draw[->,thick] (input1) -- (hidden1) node[near start, above] {$1$};
		\draw[->,thick] (input1) -- (hidden2)node[near start, below] {$-10$};

@@

		\draw[->, thick] (hidden3) -- (hidden5) node[near start, above] {{\color{red}$c$}};

@@

		\draw[->,thick] (output1) -- (output3)node[near start, above] {{\color{red}$d$}};
		\draw[->,thick] (output2) -- (output3)node[near start, below] {$1$};

	\end{tikzpicture}
	\caption{A simple example with parametric weights {\color{red}$c$} and {\color{red}$d$}}
	\label{img:FSB_example}
\end{figure}

Consider the example in Fig. \ref{img:FSB_example}. It has no bias so $s_{SR}=s_{FSB}$.
We have $\lambda(b)=d$, $\lambda(a)=\frac{cd}{2}$.
Let us do a case analysis:
The most favorable cases are $c < 0 < d$ and $d < 0 < c$: 
as $cd <0$, we have $s_{FSB}(a)=\max(0,\frac{cd}{4}) = 0$.
Now, $a$ and $\hat{a}$ need to be minimized in order to maximize $z$, because $cd <0$. 
For the triangle approximation (see Fig. \ref{triangle}), having a value $sol(a)$ for $a$, the minimal value for $\hat{a}$ is $\ReLU(sol(a))$. If we open $a$, then the value for $\hat{a}$ will also be $\ReLU(sol(a))$. That is, opening this ReLU is not helping in the approximation of $z$, as correctly predicted by the score $s_{FSB}(a)= 0$.

Case $c>0,d>0$: $s_{FSB}(a)=\frac{cd}{4}$.
The value of $a,\hat{a},b,\hat{b}$ should be maximized to maximize the value of $z$, because $W_{bz}=d>0$ and $W_{ab}=c>0$. 
%Each variation of $b$ by $\Delta(b)$ will have an impact $r(b) \cdot \Delta(b)=\frac{\Delta(b)}{2}$ on the value of $\hat{b}$, and thus $\frac{d \Delta(b)}{2}$ on $z$. 
Now, let us call $sol(a)$ the maximum value for $a$.
The maximum value for $\hat{a}$ is $r(a)\cdot (sol(a)-LB(a))$ according to the triangle approximation when $\ReLU(a)$ is not opened. Now, if $\ReLU(a)$ is opened, then 
the value of $\hat{a}$ is $\ReLU(sol(a))$: the difference is at most $r(a) \cdot LB(a) = \frac{1}{2}$. The difference $\Delta(b)$ for the value of $b$ is thus at most $\frac{c}{2}$, which means a difference at most $\frac{c}{4}$ for $\hat{b}$, using the upper function of the triangle approximation of rate $r(b)$, as the value of $\hat{b}$ should be maximized.
This means a difference of at most $\frac{cd}{4}$ on the value of $z$, which is what is computed by $s_{FSB}(a)$. Of course, the difference of opening $\ReLU(a)$ could also be 0 e.g. if $\sol(a)=\LB(a)$ or $\sol(a)=\UB(a)$.

The last case is the most problematic: 
Case $c<0,d<0$, implying that $s_{FSB}(a)=\frac{cd}{4}$, because $cd >0$.
As for the case of $c>0, d>0$, opening $\ReLU(a)$ will have an impact on the value of $\hat{a}$,
at most $\frac{1}{2}$, and hence a change $\Delta(b)$ of at most $\frac{c}{4}$ on the value of $b$.
However, as $d<0$, value of $\hat{b}$ needs to be minimized. That is, the value of $\hat{b}$ will be the value of $\ReLU(sol(b))$.
That is, the change in $\hat{b}$ will be either 0 in case $sol(b)<0$ (phase 0 of ReLU),
or $\Delta(b)<\frac{c}{2}$ otherwise (phase Id of ReLU), that is, $z$ will be modified by 
either 0 or $d \Delta(b)<\frac{cd}{2}$, as compared with $s_{FSB}(a)=\frac{cd}{4}$.

\begin{figure}[t!]
	\begin{tikzpicture}[scale=0.8, >=stealth]

		% Draw axes
		\draw[->] (-5,0) -- (4,0) node[right] {$a$};
		\draw[->] (0,-1) -- (0,3) node[above] {$\hat{a}$};

		% Draw ReLU function
		\draw[line width=0.4mm, blue] (-3,0) -- (0,0);
		\draw[thick, blue] (0,0) -- (2.5,2.5) node[below, shift={(-1.1,-1.9)}] {$\hat{a} = \ReLU(a)$};
		\draw[thick, blue] (-3,0) -- (2.5,2.5) node[above, shift={(-6.5,-2)}] {$\hat{a} = \frac{\UB}{\UB-\LB} (a-\LB)$};

@@

	\end{tikzpicture}
		\caption{Triangle approximation}
	\label{triangle}
	\end{figure}

We call {\em global rankings} (GR) such functions $s_{FSB},s_{SR}$  because they score ReLUs as accuratly as possible, considering that they do not have access to the solution $sol(a),sol(b)$ values. This analysis already hints at our novel ranking function, that will be solution-aware to be more accurate.

\iffalse
\begin{figure}[t!]
\begin{tikzpicture}[scale=1, >=stealth]

@@

\end{tikzpicture}
	\caption{}
\label{img:Utility}
\end{figure}
\fi

\iffalse
\subsection*{An example when FSB is not accurate}

@@

The plot shows that our Utility function is significantly better than FSB, and other methods.
\fi
